Clustering time series based on probability distributions across temporal granularities

### 2021-11-11 main.Rmd
finished selecting granularities

The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data. The proposed methodology aims to leverage the intrinsic data structure hidden within cyclic temporal granularities. The existing work on clustering probability distributions assumes we have an iid sample $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. <!--So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand.--> In this work, instead of considering the probability distributions of the linear time series, we assume it across different categories of any cyclic granularity. We can consider categories of an individual cyclic granularity ($A$) or combination of categories for two interacting granularities ($A*B$) to have a distribution, where $A, B$ are defined as $A = \{a_j: j=1, 2, \dots J\}$ and $B = \{b_k: k = 1, 2, \dots K\}$. For example, let us consider two cyclic granularities $A$ and $B$ representing hour-of-day and day-of-week. Then $A = \{0, 1, 2, \dots, 23\}$ and $B = \{Mon, Tue, Wed, \dots, Sun\}$. In case individual granularities are considered, there are $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. Hence clustering these customers is equivalent to clustering these collections of conditional univariate probability distributions. Towards this goal, we need to decide how to measure similarities between collections of univariate probability distributions. There are multiple ways to measure similarities depending on the aim of the analysis. <!--This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. --> This paper considers a two approaches for finding distances between time series. Both of these approaches may be useful in a practical context and, depending on the data set, may or may not propose the same customer classification. The obtained distances could be fed into a clustering algorithm to break large data sets into subgroups that can then be analyzed separately. <!--These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this.--> The methodology is explained in the Figure \ref{fig:flowchart} and each element of the pipeline is discussed.

```{r flowchart, fig.cap = "Flow chart illustrating the pipeline for methodology", out.width="100%"}
knitr::include_graphics("figs/flowchart.png")

@@

- _Find significant granularities or harmonies_

[@Gupta2021-hakear] proposes a method for choosing significant cyclic granularities and harmonies (interacting granularities that can be studied together) [@Gupta2021-gravitas], which is used in this work. We define "significant" granularities as those with significant distributional differences across categories. It is preferable to consider those since it is more probable that there will be some intriguing repeated behavior worth investigating. It should be noted that all of the observations in the study may not have the same set of important granularities. The following is a method for generating a list ($S_c$) of significant granularities for all observations:

 (a) Remove granularities from the comprehensive list that are insignificant for all observations.

 (b) select only the granularities that are significant for the majority of observations.

There will be observations in both cases where one or a few selected granularities are boring. Even in that case, having this group of observations with no interesting patterns at a granularity that regularly discovers patterns may be valuable. If, on the other hand, the granularities under examination are truly important for a group of data, unique patterns may be identified while clustering them.

>>>>

<!--The foundation of our method is unsupervised clustering algorithms based exclusively on the time-series data. The proposed methods leverage the intrinsic data structure hidden within cyclic temporal granularities.--> The existing work on clustering probability distributions assumes we have an independent and identically distributed samples $f_1(v),\dots,f_n(v)$, where $f_i(v)$ denotes the distribution from observation $i$ over some random variable $v = \{v_t: t = 0, 1, 2, \dots, T-1\}$ observed across $T$ time points. <!--So going back to the smart meter example, $f_i(v)$ is the distribution of customer $i$ and $v$ is electricity demand.--> In our approach, instead of considering the probability distributions of the linear time series, we compare them across different categories of a cyclic granularity. We can consider categories of an individual cyclic granularity ($A$) or combination of categories for two interacting granularities ($A*B$) to have a distribution, where $A, B$ are defined as $A = \{a_j: j=1, 2, \dots J\}$ and $B = \{b_k: k = 1, 2, \dots K\}$. For example, let us consider two cyclic granularities, $A$ and $B$, representing hour-of-day and day-of-week, repsectively. Then $A = \{0, 1, 2, \dots, 23\}$ and $B = \{Mon, Tue, Wed, \dots, Sun\}$. In case individual granularities are considered, there are $J = 24$ distributions of the form $f_{i,j}(v)$ or $K = 7$ distributions of the form $f_{i,k}(v)$ for each customer $i$. In case of interaction, $J*K=168$ distributions of the form $f_{i,j, k}(v)$ could be conceived for each customer $i$. Hence clustering these customers is equivalent to clustering these collections of conditional univariate probability distributions. Towards this goal, the next step is to decide how to measure similarities between collections of univariate probability distributions. <!--There are multiple ways to measure similarities depending on the aim of the analysis. <!--This paper focuses on looking at the (dis) similarity between underlying distributions that may have resulted in different patterns across different cyclic temporal granularities, that eventually have resulted in the (dis) similarity between time series. --> Here, we describe two approaches for finding distances between time series. Both of these approaches may be useful in a practical context, and produce very different but equally useful customer groupings. The distances can be supplied to any usual clustering algorithm, including $k$-means or hierarchical clustering, to group observations into a smaller more homogeneous collection. <!--These clusters may be commonly associated with real-world data segmentation. However, since the data is unlabeled a priori, more information is required to corroborate this.--> The flow of the procedures is illustrated in Figure \ref{fig:flowchart}.

```{r flowchart, fig.cap = "Flow chart illustrating the pipeline for out method for clustering time series.", out.width="100%"}
knitr::include_graphics("figs/flowchart.png")

@@

## Selecting granularities

[@Gupta2021-hakear] provides a method for determining the significance of a cyclic granularity, and a ranking of multiple cyclic granularities. (This extends to harmonies, pairs of granularities that might interact with each other.) We define "significant" granularities as those with significant distributional differences across at least one category. The reason for subsetting granularities in this way, is that, clustering algorithms perform badly in the presence of nuisance variables. Granularities that do not have some difference between categories are likely to be nuisance variables. It should be noted that all of the time series in a collection may not have the same set of significant granularities. This is the approach for managing this generate a subset ($S_c$) of significant granularities across a collection of time series:

 (a) Remove granularities from the comprehensive list that are not significant for any time series.

 (b) Select only the granularities that are significant for the majority of time series.

<!--There will be time series in both cases where one or a few selected granularities are boring. Even in that case, having this group of observations with no interesting patterns at a granularity that regularly discovers patterns may be valuable. If, on the other hand, the granularities under examination are truly important for a group of data, unique patterns may be identified while clustering them.-->
